---
title: "01_data"
format: html
---
y
# Setup

## Load Packages

```{r}
#| label: load packages
#| message: false
# load packages
## tidyverse
library(tidyverse)
library(fs)
library(httr2)
library(readxl)
library(zip)

## census
library(tidycensus)
library(tigris)
library(lehdr)

## spatial
library(sf)
library(tmap)
library(osmextract)
library(rosmium)
library(elevatr)

# create file path helpers
data_path <- dir_create("./data")
cache_path <- dir_create("./cache")

# set cache for tigris shapefiles
options(tigris_use_cache = TRUE)
tigris_cache_dir(cache_path)
```

## Set up Java Environment

```{r}
#| label: install_java
#| eval: false
rJavaEnv::java_quick_install(version = 21)
```

```{r}
#| label: check_java
rJavaEnv::java_check_version_rjava()
```

## Set Up API Keys

```{r}
#| label: setup api keys
tidycensus::census_api_key("your_key_here")
```

API Key is needed for downloading data from the US Census using {tidycensus}; see [here](https://walker-data.com/tidycensus/articles/basic-usage.html) for instructions on how to obtain an API key from the US Census Bureau.

# Define Helper Functions

Define some functions that we will use throughout the project

## Define `uncount` for {dtplyr}

Code below relies on `uncount()` from {tidyr}, which is not yet implemented in the {dtplyr} package. However, there is a workaround here: https://github.com/tidyverse/dtplyr/issues/352

```{r}
dtplyr_uncount <- function(data, weights, ..., .remove = TRUE, .id = NULL) {
  weights <- rlang::quo_squash(rlang::enquo(weights))
  
  groups <- group_vars(data)
  has_groups <- length(groups) > 0
  if (has_groups) {
    data <- ungroup(data)
  }

  vars_names <- data$vars
  if (.remove) {
    vars_names <- setdiff(vars_names, rlang::as_name(weights))
  }
  vars <- map(syms(vars_names), ~ expr(rep(!!.x, !!weights)))
  names(vars) <- vars_names
  
  if (!is.null(.id)) {
    vars <- append(vars, rlang::exprs(!!.id := sequence(!!weights)))
  }

  out <- dtplyr:::step_subset_j(
    data,
    vars = vars_names,
    j = rlang::call2(".", !!!vars)
  )
  
  if (has_groups) {
    out <- group_by(out, !!!syms(groups))
  }
  
  out
}
```

### Define ``sfc_as_cols` helper

A nice one for getting the coordinates as columns based on https://github.com/r-spatial/sf/issues/231

```{r}
sfc_as_cols <- function(x, geometry, names = c("x","y")) {
  if (missing(geometry)) {
    geometry <- sf::st_geometry(x)
  } else {
    geometry <- rlang::eval_tidy(enquo(geometry), x)
  }
  stopifnot(inherits(x,"sf") && inherits(geometry,"sfc_POINT"))
  ret <- sf::st_coordinates(geometry)
  ret <- tibble::as_tibble(ret)
  stopifnot(length(names) == ncol(ret))
  x <- x[ , !names(x) %in% names]
  ret <- setNames(ret,names)
  dplyr::bind_cols(x,ret)
}
```

# Get Census Data

First, let's get population counts at the census block group level from the 2020 Decennial Census using {tidycensus}. The code below downloads the census data, does some light cleaning by removing the water areas, and transforms it to a projected coordinate system. New York City lies within UTM Zone 18N of the North American Datum 1983 (NAD 1983) and this projection can be called through [EPSG Code:26918](https://epsg.io/26918).

## County-level Boundaries

### NYC Boroughs

NYC made up of five counties:

-   Bronx County/The Bronx (`005`)
-   Kings County/Brooklyn (`047`)
-   New York County/Manhattan (`061`)
-   Queens County/Queens (`081`)
-   Richmond County/Staten Island (`085`)

```{r}
#| label: define counties list
nyc_counties <- c(
  "061", "005", "047", "081", "085"
)
```

We can get their boundaries:

```{r}
#| label: get borough boundaries from acs
#| message: false
boundaries_boroughs <- tidycensus::get_acs(
  geography = "county",
  variables = c(total_pop = "B01001_001"),
  year = 2022,
  state = c("36"),
  county = nyc_counties,
  geometry = TRUE,
  keep_geo_vars = TRUE) |> 
  
  # project to NAD83 Zone 18N
  sf::st_transform(26918) |> 
  
  # erase water
  tigris::erase_water(area_threshold = 0.85, year = 2022)
```

We can now take a look at the boroughs by making a quick map with {tmap} (@fig-boundaries_boroughs):

```{r}
#| label: fig-boundaries_boroughs
#| fig-cap: The Five Boroughs of New York City

tm_shape(boundaries_boroughs) +
  tm_fill(
    col = "NAMELSAD",
    palette = "viridis",
    title = "The Five Boroughs \nof NYC"
  )
```

### Surrounding County Boundaries

The resulting map of the boroughs could use some surrounding counties to help with figure-ground separation. This code gets boundaries for some surrounding counties in New York, New Jersey, and Connecticut:

```{r}
#| label: get county boundaries
#| message: false
# get county boundaries
boundaries_base <- bind_rows(
  tidycensus::get_acs(
    geography = "county", 
    variables = c(total_pop = "B01001_001"),
    year = 2022,
    state = c("34"),
    county = c(003, 017, 013, 039, 023, 031), #025
    geometry = TRUE,
    cb = FALSE,
    keep_geo_vars = TRUE),
  
  tidycensus::get_acs(
    geography = "county", 
    variables = c(total_pop = "B01001_001"),
    year = 2022,
    state = c("36"),
    county = c(119, 059, 087),
    geometry = TRUE,
    cb = FALSE,
    keep_geo_vars = TRUE),
  
  tidycensus::get_acs(
    geography = "county", 
    variables = c(total_pop = "B01001_001"),
    year = 2022,
    state = c("09"),
    county = c(190),
    geometry = TRUE,
    cb = FALSE,
    keep_geo_vars = TRUE)
  ) |>
  
  # filter out water shapes
  filter(ALAND > 0) |>
  
  # project
  sf::st_transform(26918) |>
  
  # erase water area
  tigris::erase_water(area_threshold = 0.75, year = 2022)
```

We can see the improved result in @fig-boundaries_boroughs_with_base:

```{r}
#| label: fig-boundaries_boroughs_with_base
#| fig-cap: The Five Boroughs of New York City

tm_shape(boundaries_boroughs) +
  tm_fill(
    col = "NAMELSAD",
    palette = "viridis",
    title = "The Five Boroughs \nof NYC"
  ) +
  tm_shape(boundaries_base) + tm_fill(col = "grey95")
```

## Block-group Level

```{r}
#| label: get acs bg data with 2010 boundaries
#| message: false

census_bg_2010 <- tidycensus::get_acs(
  geography = "block group",
  variables = c(total_pop = "B01001_001"),
  state = "36",
  county = nyc_counties,
  year = 2019,
  output = "wide",
  geometry = TRUE,
  keep_geo_vars = TRUE
  ) |> 
  
  # rename the geoid to reflect the block group aggregation
  rename(BlockGroup_GEOID = GEOID) |>
  mutate(BlockGroup_GEOID = as.character(BlockGroup_GEOID)) |>
  
  # filter out water shapes
  filter(ALAND > 0) |>
  
  # project to NAD83 Zone 18N
  sf::st_transform(26918) |>
  
  # calculate area and set units to km^2
  mutate(bg_area_m2 = st_area(geometry),
         bg_area_km2 = units::set_units(bg_area_m2, km^2))

# get list of bg ids for filtering travel survey
census_bg_2010_id_list <- census_bg_2010 |> 
  distinct(BlockGroup_GEOID) |> 
  pull(BlockGroup_GEOID) |> 
  as.list()
```

We can now take a look at the population data by making a quick density map using {tmap}:

```{r}
#| label: fig-bg_population_density
#| fig-cap: Block Group-level Population Density
tm_shape(census_bg_2010) +
  tm_fill(
    col = "total_popE",
    area = "bg_area_km2",
    style = "fisher",
    palette = "viridis",
    title = expression("People per km"^2)
  ) +
  tm_shape(boundaries_base) + tm_fill(col = "grey95")
```

# Get MTA Travel Survey

The travel data comes from the Metropolitan Transportation Authority's (MTA) 2018 Travel Survey for New York City. For the purposes of this book, we will download linked trips and the person file:

```{r}
#| label: get mta travel survey
#| warning: false
travel_survey_path <- fs::dir_create("./data/travel_survey")

travel_survey_list <- c(
  "Linked" = "https://new.mta.info/document/29021",
  "Person" = "https://new.mta.info/document/29011",
  "Household" = "https://new.mta.info/document/29006"
)

purrr::iwalk(
  travel_survey_list, 
  ~ httr2::req_perform(
      req = httr2::request(.x) |> 
        httr2::req_cache(path = cache_path),
      path = fs::path(travel_survey_path, paste0(.y, ".xlsx")))
  )

nycts_2018_linked <- readxl::read_xlsx(
  path = fs::path(travel_survey_path, "Linked.xlsx"),
  guess_max = 3000)

nycts_2018_person <- readxl::read_xlsx(
  fs::path(travel_survey_path, "Person.xlsx"))
```

## Prepare Linked Trips

```{r}
#| label: prepare linked trips
# classify trip modes
car_mode_list <- c(
  "Auto Driver/Passenger",
  "Taxi, Car/Van Service",
  "TNC (Uber/Lyft)"
)

transit_mode_list <- c(
  "NY or MTA Bus (no subway)",
  "NYC Subway + Bus",
  "NYC Subway Only"
)

# prepare data
nycts_2018_linked <- nycts_2018_linked |> 
  # select variables of interest
  select(
    hh_id = hhid,
    person_id = personid,
    day_num = daynum,
    trip_num = tripnum,
    trip_id = tripid,
    trip_purpose = purpose,
    hh_group,
    hh_income,
    gender,
    age,
    education_level,
    driv_license = license,
    disability,
    start_lon,
    start_lat,
    stop_lon,
    stop_lat,
    orig_borough,
    dest_borough,
    orig_bg,
    dest_bg,
    home_bg,
    work_bg,
    orig_type,
    dest_type,
    departure_datetime = departure_date,
    arrival_datetime = arrival_date,
    traveldate_dow,
    mode_group_10 = mode_g10,
    modeflag_walkjogwheelchair,
    modeflag_bike,
    modeflag_citibike,
    trip_eval,
    num_transit_legs,
    access_mode = access,
    egress_mode = egress,
    subsidy_fare = commute_subsidy_1,
    subsidy_park = commute_subsidy_2,
    subsidy_toll = commute_subsidy_3,
    trip_distance = distance,
    trip_weight_wd = per_weight_wd_trips_rsadj,
    trip_weight_sat = per_weight_sat_rsadj,
    trip_weight_sun = per_weight_sun_rsadj) |> 
  
  mutate(
    # trip id to character
    trip_id = as.character(trip_id),
    hh_id = as.character(hh_id),
    gender = as_factor(gender),
    age = as_factor(age),
    education_level = as_factor(education_level),
    driv_license = as_factor(driv_license),
    disability = as_factor(disability),
    subsidy_fare = as_factor(subsidy_fare),
    subsidy_park = as_factor(subsidy_park),
    subsidy_toll = as_factor(subsidy_toll),
    
    # trip duration from survey times
    trip_duration_survey = time_length(
        difftime(arrival_datetime, departure_datetime, units = "mins"), 
      unit = "minutes"),
    
    trip_duration_survey = case_when(
      trip_duration_survey < 0 ~ 0,
      .default = trip_duration_survey),
    
    # generalize trip weight
    trip_day = case_when(
      wday(departure_datetime) == 7 ~ "sat",
      wday(departure_datetime) == 1 ~ "sun",
      .default = "wd"),
    
    trip_weight = coalesce(
      trip_weight_wd, 
      trip_weight_sat,
      trip_weight_sun),
    
    # classify trip modes
    trip_mode = case_when(
      mode_group_10 %in% car_mode_list ~ "CAR",
      mode_group_10 %in% transit_mode_list ~ "TRANSIT",
      mode_group_10 == "Walk/Bike" &
        modeflag_walkjogwheelchair == "Selected" &
        modeflag_bike != "Selected" &
        modeflag_citibike != "Selected" ~ "WALK",
      mode_group_10 == "Walk/Bike" &
        (modeflag_bike == "Selected" |
           modeflag_citibike == "Selected") ~ "BICYCLE",
      .default = "OTHER"),
    
    svyday_id = case_when(
      day_num == "First day of travel period" ~ 1,
      day_num == "Second day of travel period" ~ 2,
      day_num == "Third day of travel period" ~ 3,
      day_num == "Fourth day of travel period" ~ 4,
      day_num == "Fifth day of travel period" ~ 5,
      day_num == "Sixth day of travel period" ~ 6,
      day_num == "Seventh day of travel period" ~ 7,
      .default = 9999)
    ) |> 
  
  # filter out other modes
  #filter(trip_mode != "OTHER") |> 
  
  # order trip mode factor
  mutate(trip_mode = fct_relevel(trip_mode, "WALK", "BICYCLE", "TRANSIT", "CAR")) |> 
  select(-trip_weight_wd, -trip_weight_sat, -trip_weight_sun)
```

## Prepare Person Data
### Person Weights

Define person weights from the sample. These are sampling weights.

```{r}
nycts_person_weights <- nycts_2018_person |> 
  select(person_id = personid, starts_with("per_weight"), ends_with("complete")) |> 
  mutate(
    per_weight_day_wd = case_when(
      is.na(per_weight_wd) ~ 0,
      .default = per_weight_wd / weekdays_complete),
    
    per_weight_day_sat = case_when(
      is.na(per_weight_sat) ~ 0,
      .default = per_weight_sat / saturday_complete),
    
    per_weight_day_sun = case_when(
      is.na(per_weight_sun) ~ 0,
      .default = per_weight_sun / sunday_complete)
  ) |> 
  select(person_id, starts_with("per_weight_day"))
```

### Prepare Person Socioeconomic and Demographic Attributes

```{r}
nycts_person_seds <- nycts_2018_linked |>
  group_by(person_id) |>
  summarize(
    hh_income = first(hh_income),
    gender = first(gender),
    age = first(age),
    education_level = first(education_level),
    driv_license = first(driv_license),
    disability = first(disability),
    home_bg = first(home_bg),
    work_bg = first(work_bg)
  ) |>
  mutate(
    hh_income = fct_collapse(
      hh_income,
      "Less than $25,000" = c(
        "Less than $12,000",
        "$12,000 - $15,999",
        "$16,000- $19,999",
        "$20,000 - $24,999"
      ),
      "$25,000 - $49,999" = c(
        "$25,000 - $28,999",
        "$29,000 - $32,999",
        "$33,000 - $36,999",
        "$37,000 - $40,999",
        "$41,000 - $49,999"
      ),
      "$50,000 - $74,999" = "$50,000 - $74,999",
      "$75,000 - $99,999" = "$75,000 - $99,999",
      "$100,000 - $149,999" = "$100,000 - $149,999",
      "$150,000 or greater" = c("$150,000 - $199,999", "$200,000 - $299,999", "$300,000 or more")
    ),
    hh_income = fct_relevel(
      hh_income,
      "Less than $25,000",
      "$25,000 - $49,999",
      "$50,000 - $74,999",
      "$75,000 - $99,999",
      "$100,000 - $149,999",
      "$150,000 or greater"
    ),
    age = fct_collapse(
      age,
      "0-4 years" = "Under 5 years old",
      "5-17 years" = c("5-15 years", "16-17 years"),
      "18-34 years" = c("18-24 years", "25-29 years", "30-34 years"),
      "35-49 years" = c("35-39 years", "40-44 years", "45-49 years"),
      "50-64 years" = c("50-54 years", "55-59 years", "60-64 years"),
      "65 and older" = c(
        "65-69 years",
        "70-74 years",
        "75-79 years",
        "80-84 years",
        "85 or years older"
      )
    ),
    age = fct_relevel(
      age,
      "0-4 years",
      "5-17 years",
      "18-34 years",
      "35-49 years",
      "50-64 years",
      "65 and older"
    )
  )
```

# Get POI Data

```{r}
nyc_places_path <- fs::dir_create(fs::path(data_path, "overture_places"))
```

```{r}
#| eval: false
overture_place_themes <- read_csv("./data/overture_place_themes.csv") |> 
  select(2:1) |> 
  deframe()

study_area_bbox <- boundaries_boroughs |> 
  st_bbox()

nyc_places <- overtureR::open_curtain(type = "place", spatial_filter = study_area_bbox) |> 
  transmute(
    id,
    primary_name = names$primary,
    x = bbox$xmin,
    y = bbox$ymin,
    main_category = categories$primary,
    primary_source = sources[[1]]$dataset,
    confidence,
   geometry # currently no duckdb spatial implementation
  ) |> 
  filter(confidence > .90) |> 
  collect() |> 
  mutate(theme_primary = overture_place_themes[main_category]) |> 
  drop_na(theme_primary)

# save it to downloads so we don't keep downloading new data
save(nyc_places, file = fs::path(nyc_places_path, "nyc_places.Rdata"))
```

```{r}
load(file = fs::path(nyc_places_path, "nyc_places.Rdata"))
```

# Save

```{r}
save(
  dtplyr_uncount, 
  sfc_as_cols, 
  boundaries_boroughs,
  boundaries_base,
  census_bg_2010,
  nycts_2018_linked,
  nycts_person_weights,
  nycts_person_seds,
  nyc_places,
  file = "./output/01_data.Rdata")
```
