---
title: "02_episode_builder"
format: html
---

# Setup

Load packages

```{r}
library(tidyverse)
library(fs)
```

Load data objects:

```{r}
output_path <- fs::dir_create("./output")
load(file = fs::path(output_path, "01_data.Rdata"))
```

# Episode Builder

define episode builder as a function

```{r}
episode_builder <- function(survey_data) {
  episodes = survey_data |>
    select(
      hh_id,
      person_id,
      svyday_id,
      trip_num,
      trip_id,
      orig_bg,
      dest_bg,
      orig_type,
      dest_type,
      departure_datetime,
      arrival_datetime,
      trip_mode,
      trip_weight
    ) |>
    
    arrange(person_id, svyday_id, trip_num) |>
    
    # get a timeblock_id corresponding to continuous survey sequences, e.g. continuous activity time spread over potentially multiple days
    group_by(person_id) |>
    mutate(timeblock_id = cumsum(svyday_id - lag(svyday_id, default = first(svyday_id)) > 1) + 1) |>
    
    # replace missing orig dest types with Missing
    mutate(
      orig_type = replace_na(orig_type, "Missing"),
      dest_type = replace_na(dest_type, "Missing")
    ) |>
    ungroup() |>
    
    # add extra rows by uncounting
    mutate(count = 2) |>
    tidyr::uncount(count, .id = "segment") |>
    
    # main computations
    group_by(person_id, timeblock_id) |>
    #multidplyr::partition(cluster) |>
    
    mutate(
      # define activity classes - stays are the new second uncounted rows
      activity_class = case_when(segment == 2 ~ "Stay", .default = "Travel"),
      
      # times
      start_datetime = case_when(
        activity_class == "Stay" ~ dplyr::lag(arrival_datetime),
        .default = departure_datetime
      ),
      end_datetime = case_when(
        activity_class == "Stay" ~ dplyr::lead(departure_datetime),
        .default = arrival_datetime
      ),
      
      # locations
      start_bg = case_when(activity_class == "Stay" ~ dplyr::lag(dest_bg), .default = orig_bg),
      end_bg = case_when(activity_class == "Stay" ~ dplyr::lead(orig_bg), .default = dest_bg),
      
      # activity type
      start_activity_type = case_when(activity_class == "Stay" ~ dplyr::lag(dest_type), .default = orig_type),
      end_activity_type = case_when(activity_class == "Stay" ~ dplyr::lead(orig_type), .default = dest_type),
      
      # trip details
      trip_mode = case_when(activity_class == "Stay" ~ NA_character_, .default = trip_mode),
      trip_id = case_when(activity_class == "Stay" ~ NA_character_, .default = trip_id),
      trip_weight = case_when(activity_class == "Stay" ~ NA, .default = trip_weight)
    ) |>
    
    select(-segment) |>
    
    # LAST STAYS in timeblock_id sequence
    ## fix each person's last stay episodes where leads will produce NAs
    mutate(
      end_datetime = case_when(
        is.na(end_datetime) ~ lubridate::ymd_hms(paste0(
          lubridate::date(start_datetime), " 23:59:59"
        )),
        .default = end_datetime
      ),
      end_bg = case_when(is.na(end_bg) ~ start_bg, .default = end_bg),
      end_activity_type = case_when(is.na(end_activity_type) ~ "Last Stay", .default = end_activity_type)
    ) |>
    
    # INITIAL STAYS in timeblock_id sequence
    ## add a new row for the first day stay for a person before any trip occurs
    slice(1, row_number()) |>
    
    ## fix up the first row only
    mutate(
      ## fix the start datetime to midnight
      start_datetime = case_when(
        row_number() == 1 ~ lubridate::floor_date(lead(start_datetime), unit = "day"), 
        .default = start_datetime),
      
      ## classify initial stays (don't want to assume an activity)
      start_activity_type = case_when(
        row_number() == 1 ~ "Initial Stay", 
        .default = start_activity_type),
      
      # pull end info from next row's start
      end_datetime = case_when(row_number() == 1 ~ lead(start_datetime), .default = end_datetime),
      end_bg = case_when(row_number() == 1 ~ lead(start_bg), .default = end_bg),
      end_activity_type = case_when(row_number() == 1 ~ lead(start_activity_type), .default = end_activity_type),
      
      ## fix trip details to NA as this is a stay
      trip_mode = case_when(row_number() == 1 ~ NA_character_, .default = trip_mode),
      trip_id = case_when(row_number() == 1 ~ NA_character_, .default = trip_id),
      trip_weight = case_when(row_number() == 1 ~ NA, .default = trip_weight),
      
      ## fix activity info
      activity_class = case_when(row_number() == 1 ~ "Stay", .default = activity_class),
      activity_desc = case_when(
        activity_class == "Stay" & row_number() == n() ~ end_activity_type,
        activity_class == "Stay" &
          row_number() != n() ~ start_activity_type,
        
        .default = stringr::str_to_title(trip_mode)
      )
    ) |>
    
    # finish up
    # add activity duration
    mutate(
      activity_duration = difftime(end_datetime, start_datetime, units = "mins"),
      activity_duration = case_when(activity_duration < 0 ~ make_difftime(0, "minutes"), .default = activity_duration)
    ) |>
    
    # do some calendar day-based computations
    group_by(person_id, svyday_id) |>
    
    # add proper 24h calendar day id and standard weekday
    mutate(episode_id = row_number()) |>
    
    # add standard weekday, if a final episode of the day happened past midnight, this keeps the start day intact
    mutate(std_wday = case_when(
      wday(first(start_datetime)) == 1 ~ "sunday",
      wday(first(start_datetime)) == 7 ~ "saturday",
      .default = "weekday"
    )) |> 
    
    ungroup() |>
    
    # clean up
    select(
      hh_id,
      person_id,
      svyday_id,
      timeblock_id,
      std_wday,
      episode_id,
      trip_id,
      trip_mode,
      trip_weight,
      activity_class,
      activity_desc,
      activity_duration,
      start_datetime,
      end_datetime,
      start_activity_type,
      end_activity_type,
      start_bg,
      end_bg
    )
  
  return(episodes)
}
```

Build the episodes (this takes a few minutes to run):

```{r}
episodes <- episode_builder(nycts_2018_linked)
```

## Person Weights

add person weights

```{r}
episodes <- episodes |>
  left_join(nycts_person_weights, by = "person_id") |>
  group_by(person_id, svyday_id) |>
  mutate(
    # get person weight by survey standard day
    per_weight = case_when(
      std_wday == "weekday" ~ per_weight_day_wd,
      std_wday == "saturday" ~ per_weight_day_sat,
      .default = per_weight_day_sun
    ),
    
    # if person weight is zero (error in survey data), take the mean trip weight instead (very few examples)
    per_weight = case_when(per_weight == 0 ~ mean(trip_weight, na.rm = TRUE), .default = per_weight)
  ) |>
  ungroup() 
```

## Do some quality control

- valid stay rules
- starts and ends at same Block Group`valid_path`:

valid cadence - search for breaks in the activity sequence, for example ID .. ends a day at home but then starts the next day at a different destinations, so a trip is implied/missing which breaks the sequence

```{r}
episodes <- episodes |>
  group_by(person_id) |>
  
  # FLAG INVALID STAYS AND PATHS
  mutate(
    ## find where any start_bg does not match end_bg of previous episode - break in sequence
    valid_path_line = case_when(
      activity_desc == "Initial Stay" ~ "yes",
      activity_desc != "Initial Stay" &
        start_bg == lag(end_bg) ~ "yes",
      .default = "no"
    ),
    
    ## find where any stay takes place across two bgs (implying travel and a missed trip)
    valid_stay_location = case_when(
      activity_class == "Stay" & start_bg == end_bg ~ "yes",
      activity_class == "Travel" ~ "travel",
      .default = "no"
    ),
    
    ## is there a valid stay-trip sequence or has a missing/implied trip occurred?
    valid_stay_trip_sequence = case_when(
      activity_class == "Stay" &
        (activity_desc != "Initial Stay" |
           activity_desc != "Last Stay") &
        start_activity_type != end_activity_type ~ "no",
      .default = "yes"
    ),
    
    ## is there a break in the sequence of survey days - a day missed? will mess with stay durations
    valid_svyday_sequence = case_when(
      difftime(end_datetime, start_datetime, units = "days") > 1 ~ "no",
      .default = "yes"
    ),
  ) |>
  ungroup() |>
  group_by(
    activity_class,
    activity_desc,
    valid_stay_location,
    valid_stay_trip_sequence,
    valid_svyday_sequence
  ) |>
  mutate(
    duration_pctile = quantile(activity_duration, 0.99),
    valid_duration = case_when(activity_duration > duration_pctile ~ "no", .default = "yes")
  ) |>
  ungroup() |>
  select(-duration_pctile)
```

# Make Tempograms

```{r}
activites_list <- episodes |> 
  ungroup() |> 
  distinct(activity_desc) |> 
  pull(activity_desc) |> 
  as.list()

activites_short_names <- c(
  "Initial Stay" = "Initial Stay",
  "Transit" = "Transit",
  "Work" = "Work",
  "Walk" = "Walk",
  "Home" = "Home",
  "Restaurant (dine-in or take-out)" = "Restaurant",
  "Friend/Family member’s house" = "Friend/Family",
  "Last Stay" = "Last Stay",
  "K-12 School or daycare" = "Education",
  "Shopping" = "Shopping",
  "Recreation/Entertainment/Exercise (e.g., movies, theater, park)" = "Recreation",
  "Errands or personal business (e.g. bank, post office)" = "Errands",
  "Other" = "Other",
  "Car" = "Car",
  "Work-related place (e.g., meeting, second job, delivery)" = "Work-related",
  "Missing" = "Missing",
  "Religious (e.g., church, synagogue)" = "Religious",
  "Airport" = "Airport",
  "Doctor/Medical Service/Hospital" = "Medical",
  "College/University" = "University",
  "Bicycle" = "Bicycle"
)

activites_short_names_summary <- c(
  "Initial Stay" = "Initial Stay",
  "Transit" = "Travel",
  "Work" = "Work",
  "Walk" = "Travel",
  "Home" = "Home",
  "Restaurant (dine-in or take-out)" = "Restaurant",
  "Friend/Family member’s house" = "Social",
  "Last Stay" = "Last Stay",
  "K-12 School or daycare" = "Edu: K-12",
  "Shopping" = "Shopping",
  "Recreation/Entertainment/Exercise (e.g., movies, theater, park)" = "Recreation",
  "Errands or personal business (e.g. bank, post office)" = "Errands",
  "Other" = "Other",
  "Car" = "Travel",
  "Work-related place (e.g., meeting, second job, delivery)" = "Work",
  "Missing" = "Missing",
  "Religious (e.g., church, synagogue)" = "Religious",
  "Airport" = "Travel",
  "Doctor/Medical Service/Hospital" = "Errands",
  "College/University" = "Edu: College",
  "Bicycle" = "Travel"
) |> 
  enframe() |> 
  deframe()
```

```{r}
time_unit = 15

activity_exclude_list <- c("Other", "Missing", "Initial Stay", "Last Stay")

episodes_tempogram <- episodes |>
  ungroup() |>
  group_by(person_id, timeblock_id) |>
  mutate(
    start_datetime_r = round_date(start_datetime, paste0(15, " minutes")),
    activity_duration_r = round(as.numeric(activity_duration) / 15) * 15,
    duration_multiple = activity_duration_r / 15,
  ) |>
  collect() |>
  ungroup() |>
  tidyr::uncount(duration_multiple, .id = "uncount_n") |>
  mutate(
    activity_desc_sum = activites_short_names_summary[activity_desc],
    activity_time = time_unit,
    activity_time_w = per_weight * time_unit,
    cum_time = time_unit * (uncount_n - 1),
    cum_start_datetime = start_datetime_r + minutes(cum_time),
  ) |>
  filter(!activity_desc_sum %in% activity_exclude_list) |>
  group_by(activity_desc_sum,
           std_wday,
           std_time = hms::as_hms(cum_start_datetime)) |>
  summarize(
    activity_time = sum(activity_time),
    activity_time_w = sum(activity_time_w)
  ) |>
  ungroup() |>
  mutate(std_wday = fct_relevel(std_wday, "weekday", "saturday", "sunday"), )
```

```{r}
#| fig-height: 8
#| fig-width: 8
episodes_tempogram |>
  ggplot(aes(x = std_time, y = activity_time_w, fill = activity_desc_sum, colour = activity_desc_sum)) +
  geom_area(position = position_fill(), alpha = .8) +
  scale_x_time(breaks = scales::date_breaks("3 hours")) +
  scale_y_continuous(labels = scales::percent) +
  facet_wrap(vars(std_wday), nrow = 3) +
  theme_minimal() +
  theme(
    legend.position = "right",
    axis.title.x = element_blank(),
    axis.title.y = element_blank(),
    legend.title = element_blank()
  )
```

# Save 

```{r}
save(
  episode_builder, 
  episodes, 
  episodes_tempogram, 
  file = fs::path(output_path, "02_episodes.Rdata"))
```
