---
title: "04_networks"
format: html
---

# Setup

```{r}
#| label: setup
options(java.parameters = "-Xmx64G")

library(tidyverse)
library(fs)
library(sf)
library(tmap)

library(r5r)
library(tidytransit)

library(furrr)
plan(multisession, workers = parallel::detectCores() - 2)
```

```{r}
# create data directories
cache_path <- dir_create("./cache")

data_path <- dir_create("./data")

elevation_path <- fs::dir_create(fs::path(data_path, "elevation_15m"))
osm_data_path <- fs::dir_create(data_path, "osm")
gtfs_2018_path <- fs::dir_create(fs::path(data_path, "gtfs_2018"))
```

# Download Data

# Get GTFS Files

Downloading current and historical GTFS files for different transit providers from [Transitland](https://www.transit.land). See [here](https://www.interline.io/transitland/plans-pricing/) for more information on how to obtain a key. Transitland's basic free tier allows for the downloading of current feed versions, while a separate hobbyist/academic free tier enables a user to download historical feeds.

```{r}
transit_land_api_key <- "your_key_here"
```

## GTFS for Spring 2018

The MTA Travel Survey downloaded below was fielded from March-June 2018. To calculate trip flows on the network, the code below obtains GTFS files that cover this time period:

```{r}
#| label: get gtfs for 2018
#| eval: false

gtfs_2018_list <- c(
  "nyct_subway" = "174e58c5a32eb4404b0358ce8dbf179dfe914572",
  
  "nyct_bus_bronx_1" = "7dbe46daa61f9a996476649e6a141ec306dd4796",
  "nyct_bus_bronx_2" = "e4be1ad7a2372b25c1c6ec6f294e359d985849ed",
  
  "nyct_bus_manhattan_1" = "1da101485fa5bab9e8d504c898485e114a1eb5d6",
  "nyct_bus_manhattan_2" = "339d3b2ff1a2a17495d0f36f478e82a3cc2fe720",
  
  "nyct_bus_brooklyn_1" = "c31801687408edecdd2b42a86378340e5fdfb14f",
  "nyct_bus_brooklyn_2" = "f458cf2e9c3dbd698f8cc1c1ea4db664a11cb6b4",
  
  "nyct_bus_staten_1" = "b8ea7709196b1445f5b07114abeb1b3ff1bc369e",
  "nyct_bus_staten_2" = "50735207cf3b244b6158cb38b7530d82a8ba6b4e",
  
  "nyct_bus_queens_1" = "f9d88f3d17844b4c2cbd10e5e47574c698eecaa7",
  "nyct_bus_queens_2" = "246bad93b71de78b0e789327d7a74e201843ea90",
  
  "nyc_busco_1" = "2010595e7244adf85c81c9ee6e3f2c4445e2aca1",
  "nyc_busco_2" = "a6dc342ba045f320e4e98aaec6fc5ee57a046fdf",
  
  "si_ferry" = "55cb4f102021db7ab4ffd74b8a734e9c386cf544",
  "nyc_ferry" = "dbc75610cfd814d5bd61edafbf442c1d5d990458"
)

purrr::iwalk(
  gtfs_2018_list, 
  ~ httr2::req_perform(
      req = httr2::request(paste0("https://transit.land/api/v2/rest/feed_versions/", .x, "/download?apikey=", transit_land_api_key)) |> 
        httr2::req_cache(path = cache_path),
      path = fs::path(gtfs_2018_path, paste0(.y, ".zip")))
  )
```

Unfortunately, the Staten Island Ferry GTFS file has an error: the required `route_short_name` column is named `short_name` in the `routes.txt` file. The code below unzips the archive, reads in the file, renames the column, writes it back out, deletes the old archive, and makes a new `.zip` with the fixed file:

```{r}
#| label: fix si_ferry gtfs
#| message: false
#| eval: false

# unzip staten island ferry gtfs
zip::unzip(
  zipfile = fs::path(gtfs_2018_path, "si_ferry.zip"),
  exdir = fs::path(gtfs_2018_path, "si_ferry"))

# read in routes, change name, and rewrite to txt
read_delim(
    file = fs::path(gtfs_2018_path, "si_ferry", "routes.txt"),
    delim = ",") |> 
  rename(route_short_name = short_name) |> 
  write_delim( 
    file = fs::path(gtfs_2018_path, "si_ferry", "routes.txt"), 
    delim = ",")

# delete old zip
fs::file_delete(fs::path(gtfs_2018_path, "si_ferry.zip"))

# make new zip
zip::zip(
  zipfile = fs::path(gtfs_2018_path, "si_ferry.zip"), 
  files = fs::dir_ls(fs::path(gtfs_2018_path, "si_ferry")),
  mode = "cherry-pick")

# delete working folder
fs::file_delete(fs::path(gtfs_2018_path, "si_ferry"))
```

# Get OSM Data

Next, we will get data from OpenStreetMap for the New York City metro region using the {osmextract} [@gilardi2022] package. The `openstreetmap_fr` provider has an extract for New York City:

```{r}
#| label: check osmextract for nyc
osmextract::oe_match("new york city", provider = "openstreetmap_fr")
```

This data can then be downloaded using the code below. However, this downloads the most current version of the extract. Because there is no easy way to download historic OSM data, an OSM extract from November 2023 that aligns with the GTFS data is stored in the `data/osm_2023` folder.

```{r}
#| label: get osm for nyc
#| eval: false
osmextract::oe_get(
  place = "new york city", provider = "openstreetmap_fr",
  download_directory = osm_data_path,
  download_only = TRUE,
  skip_vectortranslate = TRUE
)
```

# Get Elevation Data

An elevation raster for the study area can be downloaded using the {elevatr} package. In this case, we will make use of the elevation tiles from Amazon Web Service (AWS) at around a 15m raster resolution, clipped to the bounding box that contains the `boundaries_boroughs`:

```{r}
#| label: get elevation raster from AWS
#| message: false
#| eval: false

elevation_raster <- 
  elevatr::get_elev_raster(
    locations = boundaries_boroughs, 
    z = 12, # 13 = about 7m; 12 = about 15m
    src = "aws", 
    clip = "bbox", 
    expand = 1000) |> 
  
  # convert to a terra raster
  terra::rast()
```

```{r}
#| label: save raster as tif
#| eval: false

terra::writeRaster(
  x = elevation_raster,
  filename = fs::path(elevation_path, "nyc_elevation_15m.tif"),
  overwrite = TRUE)
```

# Forward Network

```{r}
#| label: copy r5 inputs for forward time network
#| eval: false

r5_path_forward <- dir_create("./r5_graph_forward") # for the r5 network graph in forward time direction

r5_forward_inputs_list <- c(
  fs::path(osm_data_path, "openstreetmap_fr_new_york_city-latest.osm.pbf"),
  fs::dir_ls(gtfs_2018_path, regexp = "*.zip"),
  fs::path(elevation_path, "nyc_elevation_15m.tif")
)

file_copy(
  path = r5_forward_inputs_list,
  new_path = r5_path_forward,
  overwrite = TRUE
)
```

## Build Network

```{r}
#| eval: false
r5_graph_forward <- r5r::setup_r5(data_path = r5_path_forward)
```

# Reverse Transit Network
## Copy Street Network and Elevation

```{r}
#| label: copy r5 inputs for reverse time network
r5_path_reverse <- fs::dir_create("./r5_graph_reverse") # for the r5 network graph in reverse time direction

r5_reverse_inputs_list <- c(
  fs::path(osm_data_path, "openstreetmap_fr_new_york_city-latest.osm.pbf"),
  fs::path(elevation_path, "nyc_elevation_15m.tif")
)

file_copy(
  path = r5_reverse_inputs_list,
  new_path = r5_path_reverse,
  overwrite = TRUE
)
```

## Get GTFS Files and Reverse Time
### Define Time Reversal Function

```{r}
#| label: define reverse time and datetime functions
# define reverse time function
reverse_time <- function(forward_time){
  hh_rev = str_pad(abs(hour(forward_time) - 23), width = 2, side = "left", pad = "0")
  mm_rev = str_pad(abs(minute(forward_time) - 59), width = 2, side = "left", pad = "0")
  ss_rev = str_pad(abs(second(forward_time) - 59), width = 2, side = "left", pad = "0")
  time_rev = paste(hh_rev, mm_rev, ss_rev, sep = ":")
  return(time_rev)
}

# define reverse datetime function
reverse_datetime <- function(forward_datetime){
  forward_date = as_date(forward_datetime)
  time_rev = reverse_time(forward_datetime)
  datetime_rev = ymd_hms(paste(forward_date, time_rev, sep = " "))
  return(datetime_rev)
}
```

### Define GTFS Time Reversal Function

```{r}
#| label: define gtfs reverse time function
gtfs_reverse_time <- function(gtfs_zip, gtfs_work_folder, output_folder){
  z_fullname = fs::path_file(gtfs_zip)
  z_name = fs::path_ext_remove(z_fullname)
  z_path = fs::path_dir(gtfs_zip)
  z_out_fullname = paste0(z_name, "_rev.zip")
  
  work_path = fs::dir_create(fs::path(gtfs_work_folder, z_name))
  #z_tmp_dir <- tempdir()
  
  # unzip gtfs folder
  zip::unzip(zipfile = gtfs_zip, exdir = work_path)
  
  # read stop_times
  stop_times = readr::read_delim(file = fs::path(work_path, "stop_times.txt"), col_types = cols(.default = "c")) |> 
    
    # reverse stop times
    mutate(arrival_time_rev = reverse_time(lubridate::hms(arrival_time)),
           departure_time_rev = reverse_time(lubridate::hms(departure_time))) |> 
    
    # reverse stop sequence and re-sort
    group_by(trip_id) |> 
    mutate(stop_sequence_rev = abs(as.numeric(stop_sequence) - (max(as.numeric(stop_sequence)))-1)) |> 
    arrange(stop_sequence_rev, .by_group = TRUE) |> 
    ungroup() |> 
    
    # finish up
    mutate(arrival_time = departure_time_rev,
           departure_time = arrival_time_rev,
           stop_sequence = stop_sequence_rev) |>
    dplyr::select(-arrival_time_rev, -departure_time_rev, -stop_sequence_rev)
  
  # reverse stop sequence for shapes
  shapes = readr::read_delim(file = fs::path(work_path, "shapes.txt"), col_types = cols(.default = "c")) |>
    group_by(shape_id) |>
    mutate(shape_pt_sequence_rev = abs(as.numeric(shape_pt_sequence) - (max(as.numeric(shape_pt_sequence))))) |>
    arrange(shape_pt_sequence_rev, .by_group = TRUE) |>
    ungroup() |>
  
    # finish up
    mutate(shape_pt_sequence = shape_pt_sequence_rev) |>
    dplyr::select(-shape_pt_sequence_rev)
  
  # write output
  write_delim(stop_times, file = file.path(work_path, "stop_times.txt"), na = "", delim = ",")
  write_delim(shapes, file = file.path(work_path, "shapes.txt"), na = "", delim = ",")
  
  # create new zip
  zip::zip(zipfile = path(output_folder, z_out_fullname), 
           files = dir_ls(work_path), 
           mode = "cherry-pick")
}
```

```{r}
gtfs_work_path <- fs::dir_create("./gtfs_work_folder")

# get a list of gtfs zip files in the gtfs directory
gtfs_zip_list <- dir_ls(r5_path_forward, regexp = "*.zip")

# reverse time in the gtfs files and send them to the reverse time network folder
furrr::future_walk(
  gtfs_zip_list,
  ~ gtfs_reverse_time(
    gtfs_zip = .x,
    gtfs_work_folder = gtfs_work_path,
    output_folder = r5_path_reverse
  )
)
```

## Build Network

```{r}
#| eval: false
r5_graph_reverse <- setup_r5(data_path = r5_path_reverse, verbose = FALSE)
```
