---
title: "08_accessibility"
format: html
---

```{r}
#| label: setup
options(java.parameters = "-Xmx64G")

library(tidyverse)
library(patchwork)
library(dtplyr)
library(fs)
library(sf)
library(stars)
library(terra)
library(tmap)
library(r5r)
library(arrow)
```

# Setup

```{r}
output_path <- fs::dir_create("./output")
load(file = fs::path(output_path, "01_data.Rdata"))
load(file = fs::path(output_path, "03_st_paths.Rdata"))
```

## prepare study area grid

```{r}
cell_size = 100

study_area_bbox <- boundaries_boroughs |> 
  #filter(NAMELSAD == "New York County") |> 
  #st_buffer(1000) |> 
  st_bbox()

study_area_boundaries <- boundaries_boroughs |> 
  st_crop(study_area_bbox) |> 
  st_collection_extract("POLYGON")

study_area_raster <- boundaries_boroughs |>
  select(geometry) |> 
  st_rasterize(
    template = st_as_stars(
      study_area_bbox,
      dx = cell_size,
      dy = cell_size,
      pretty = TRUE,
      values = NA_real_))

study_area_points <- study_area_raster |> 
  st_xy2sfc(as_points = TRUE) |> 
  st_as_sf() |> 
  select(-ID) |> 
  mutate(id = as.character(row_number()))

study_area_polygons <- study_area_raster |> 
  st_xy2sfc(as_points = FALSE) |> 
  st_as_sf() |> 
  select(-ID) |> 
  st_join(study_area_points)
```

# Get Trip Paths

## Get Flex Trips

```{r}
flex_activities <- c(
  "Restaurant (dine-in or take-out)",
  "Recreation/Entertainment/Exercise (e.g., movies, theater, park)",
  "Shopping"
)

flex_trips <- st_paths |> 
  drop_na(trip_id) |> 
  filter(end_activity_type %in% flex_activities) |> 
  sf::st_zm(drop = TRUE) |> 
  mutate(
    start_point = lwgeom::st_startpoint(geometry),
    end_point = lwgeom::st_endpoint(geometry))

trip_weights <- flex_trips |> 
  st_drop_geometry() |> 
  select(trip_id, trip_weight) |> 
  mutate(trip_id = as.character(trip_id))
```

## calculate trip paths

```{r}
r5_path_forward <- dir_create("./r5_graph_forward")
r5_graph_forward <- r5r::build_network(data_path = r5_path_forward)
```

```{r}
trip_paths_path <- fs::dir_create("./trip_paths_flex")
```

```{r}
#| eval: false

trip_paths <- flex_trips |> 
  filter(trip_mode != "OTHER") |>
  #mutate(from_id = trip_id, to_id = trip_id) |>
  # unify start_datetime for non-transit modes
  mutate(
    std_wday = case_when(
      wday((start_datetime)) == 1 ~ "sunday",
      wday((start_datetime)) == 7 ~ "saturday",
      TRUE ~ "weekday")) |> 
  filter(std_wday == "weekday") |> 
  mutate(
    #start_datetime = case_when(trip_mode != "TRANSIT" ~ NA, TRUE ~ start_datetime),
    std_day = case_when(
      # sunday
      wday(start_datetime) == 1 ~ "2018-03-25",
      # saturday
      wday(start_datetime) == 7 ~ "2018-03-24",
      # weekday
      .default = "2018-03-23" 
    ),
    std_departure_datetime = case_when(
      trip_mode == "TRANSIT" ~ ymd_hms(paste0(std_day, " ", hms::as_hms(start_datetime))), # keep original start time for transit with standard weekday
      TRUE ~ ymd_hms(paste0(std_day, " 08:00:00")) # all other modes default to 8:00AM (doesn't matter for routing but helps batching)
    ),
    std_departure_datetime = round_date(std_departure_datetime, "5 minutes") # round to nearest 5 min for batching
  ) |> 
  group_by(trip_mode, std_departure_datetime) |> 
  group_map(
    ~ r5r::detailed_itineraries(
      r5r_network = r5_graph_forward,
      origins = .x |> select(id = trip_id, start_point) |> st_set_geometry("start_point") |> st_transform(crs = 4326),
      destinations = .x |> select(id = trip_id, end_point) |> st_set_geometry("end_point") |> st_transform(crs = 4326),
      mode = .y |> pull("trip_mode"),
      departure_datetime = .y |> pull("std_departure_datetime")
    ) |> 
      add_column(.y)) |> 
  bind_rows() |> 
  st_transform(crs = 26918) |> 
  left_join(trip_weights, by = c("from_id" = "trip_id")) 

save(trip_paths, file = fs::path(trip_paths_path, "trip_paths.Rdata")) 
```

# Get Impedance Functions from Trip Paths

```{r}
#| label: load trip paths
load(file = fs::path(trip_paths_path, "trip_paths.Rdata"))
```

```{r}
flex_trips <- trip_paths |> 
  st_drop_geometry() |> 
  mutate(
    std_wday = case_when(
      wday((std_departure_datetime)) == 1 ~ "sunday",
      wday((std_departure_datetime)) == 7 ~ "saturday",
      TRUE ~ "weekday")) |> 
  group_by(trip_id = from_id, trip_mode) |> 
  summarize(count = n(),
            trip_weight = first(trip_weight),
            trip_duration = first(total_duration)) |> 
  group_by(trip_mode) |> 
  mutate(trip_weight_prop = trip_weight / sum(trip_weight)) |> 
  ungroup()
```

Get CDFs by mode, using trip weights:

```{r}
modal_cdfs <- flex_trips |> 
  group_by(trip_mode) |> 
  nest() |> 
  mutate(
    pdf = map(data, 
      ~ density(x = .x |>  pull("trip_duration"), 
                weights = .x |>  pull("trip_weight_prop"),
                from = 0, 
                to = 180,
                adjust = 2)),
    cdf = map(pdf, 
      ~spatstat.univar::CDF(f = .x))) |> 
  select(trip_mode, cdf) |> 
  deframe()

impedance_cdf <- list(
  BICYCLE = function(t_ij){1-modal_cdfs$BICYCLE(t_ij)},
  CAR = function(t_ij){1-modal_cdfs$CAR(t_ij)},
  TRANSIT = function(t_ij){1-modal_cdfs$TRANSIT(t_ij)},
  WALK = function(t_ij){1-modal_cdfs$WALK(t_ij)}
)

modal_cdf_df <-
  data.frame(t_ij = seq(from = 0, to = 120, by = 1)) %>%
  mutate(across(t_ij, .fns = impedance_cdf, .names = "cdf_{fn}")) %>%
  pivot_longer(
    cols = starts_with("cdf"),
    names_to = "trip_mode",
    values_to = "impedance_weight",
    names_prefix = "cdf_"
  ) %>%
  mutate(trip_mode = fct_relevel(
    trip_mode,
    "WALK", "BICYCLE", "TRANSIT", "CAR"))
```

Make plots:

```{r}
pdf_plot <- flex_trips |> 
  #ggplot(aes(x = total_duration, colour = trip_mode)) + 
  #geom_density(adjust = 2)
  ggplot(aes(x = trip_duration,
      colour = trip_mode,
      #fill = trip_mode,
      weight = trip_weight_prop)) +
  geom_density(adjust = 2,
               alpha = .1,
               lwd = 1) +
  viridis::scale_color_viridis(discrete = TRUE, option = "D", end = .97) +
  #viridis::scale_fill_viridis(discrete = TRUE, begin = .2) +
  scale_x_continuous(limits = c(0, 90), 
                     breaks = seq(0, 90, by = 30)) +
  ylab("probability density") +
  theme_minimal() +
  guides(fill="none", colour = "none") +
  theme(axis.title.x=element_blank())

cdf_plot <- ggplot(modal_cdf_df, 
       aes(x = t_ij, 
           y = impedance_weight, 
           colour = trip_mode)) +
  geom_line(size = 1) +
  scale_x_continuous(limits = c(0, 90), 
                     breaks = seq(0, 90, by = 30)) +
  viridis::scale_color_viridis(discrete = TRUE, option = "D", end = .97) +
  guides(color=guide_legend(title="trip mode")) +
  xlab("travel time (minutes)") +    
  ylab("1 - cumulative probability") + 
  theme_minimal()

pdf_cdf_plot <- pdf_plot / cdf_plot + 
  plot_layout(guides = "collect") & 
  theme(legend.position = "right")
pdf_cdf_plot
```

# Get Opportunities by Cell

```{r}
overture_flexthemes <- c(
  "active_life",
  "arts_and_entertainment",
  "attractions_and_activities",
  "eat_and_drink",
  "retail"
)

# count opportunities by cell
nyc_places_flex <- nyc_places |> 
  rename(poi_id = id) |> 
  filter(theme_primary %in% overture_flexthemes) |> 
  st_transform(crs = 26918) |> 
  st_join(study_area_polygons) |>
  drop_na(id) |> 
  st_drop_geometry() |> 
  group_by(id) |> 
  summarize(opp_weight = n())

# join back to study area points
study_area_points <- study_area_points |> 
  left_join(nyc_places_flex, by = "id") |> 
  mutate(opp_weight = replace_na(opp_weight, 0)) |> 
  st_join(census_bg_2010 |> select(block_group = BlockGroup_GEOID, county = COUNTYFP, geometry)) |> 
  mutate(
    block_group = replace_na(block_group, "9999"),
    county = replace_na(county, "9999"))

opportunities_raster <- study_area_points |> 
  select(-id) |> 
  filter(opp_weight > 0) |> 
  st_as_stars() |> 
  st_sfc2xy()
```

# Calculate Travel Time Matrix

calculate ttm using BGs as origins, assume PM peak for weekday travel

```{r}
#| eval: false
#| warning: false
trip_mode <- c("walk", "bicycle", "transit", "car") 
ttms_path <- fs::dir_create("./ttms_100m")
  
census_bg_2010_origins <- census_bg_2010 |>
  select(id = BlockGroup_GEOID, county = COUNTYFP, total_popE, geometry) |> 
  st_centroid()

census_bg_2010_origins |> 
  crossing(trip_mode) |>
  st_as_sf() |> 
  group_by(trip_mode, county) |>
  group_walk(
    ~ r5r::travel_time_matrix(
      r5r_network = r5_graph_forward,
      origins = .x |> st_transform(crs = 4326),
      destinations = study_area_points |> st_transform(crs = 4326),
      mode = .y |> pull(trip_mode),
      departure_datetime = ymd_hms("2018-03-23 16:00:00"),
      max_trip_duration = 90) |> 
      add_column(.y) |>
      left_join(study_area_points |> st_drop_geometry() |> select(id, opp_weight), by = c("to_id" = "id")) |> 
      left_join(census_bg_2010_origins |> st_drop_geometry() |> select(id, total_popE), by = c("from_id" = "id")) |> 
      group_by(trip_mode, county) |> 
      arrow::write_dataset(ttms_path)
  )
```

# Calculate Accessibility
## Register impedance functions into {arrow}

```{r}
#| eval: true
register_scalar_function(
  name = "impedance_f_walk",
  fun = function(context, t_ij) {impedance_cdf$WALK(t_ij)},
  in_type = schema(t_ij = int32()),
  out_type = float64(),
  auto_convert = TRUE
)

register_scalar_function(
  name = "impedance_f_bicycle",
  fun = function(context, t_ij) {impedance_cdf$BICYCLE(t_ij)},
  in_type = schema(t_ij = int32()),
  out_type = float64(),
  auto_convert = TRUE
)

register_scalar_function(
  name = "impedance_f_transit",
  fun = function(context, t_ij) {impedance_cdf$TRANSIT(t_ij)},
  in_type = schema(t_ij = int32()),
  out_type = float64(),
  auto_convert = TRUE
)

register_scalar_function(
  name = "impedance_f_car",
  fun = function(context, t_ij) {impedance_cdf$CAR(t_ij)},
  in_type = schema(t_ij = int32()),
  out_type = float64(),
  auto_convert = TRUE
)
```

## Calculate Market Potential

```{r}
place_marketpotential_by_mode <- arrow::open_dataset("./ttms_100m") |>
  mutate(
    grav_weight = case_when(
      trip_mode == "walk" ~ impedance_f_walk(travel_time_p50),
      trip_mode == "bicycle" ~ impedance_f_bicycle(travel_time_p50),
      trip_mode == "transit" ~ impedance_f_transit(travel_time_p50),
      TRUE ~ impedance_f_car(travel_time_p50),
    ),
    grav_markpop = total_popE * grav_weight
  ) |>
  group_by(to_id, trip_mode) |>
  summarize(grav_markpop = sum(grav_markpop)) |> 
  collect()

```

Make into a {stars} object:

```{r}
place_marketpotential_by_mode_weekday_stars <- study_area_points |>
  left_join(place_marketpotential_by_mode, by = c("id" = "to_id")) |>
  drop_na(trip_mode) |>
  mutate(
    trip_mode = fct_relevel(trip_mode, "walk", "bicycle", "transit", "car")
  ) |> 
  sfc_as_cols() |>
  st_drop_geometry() |>
  st_as_stars(dims = c("x", "y", "trip_mode")) |>
  st_set_crs(26918)
```

# Save

```{r}
save(
  impedance_cdf,
  pdf_cdf_plot,
  place_marketpotential_by_mode_weekday_stars,
  file = "./output/07_place_access.Rdata"
)
```
