---
title: "Untitled"
format: html
---


```{r}
options(java.parameters = "-Xmx96G")

library(tidyverse)
library(dtplyr)
library(fs)
library(sf)
library(sfheaders)
library(stars)
library(terra)
library(tmap)

library(tinytable)

library(r5r)

## Load Data
data_path <- fs::dir_create("./data")
output_path <- fs::dir_create("./output")

load(file = "./output/01_data.Rdata")
load(file = "./output/02_episodes.Rdata")
load(file = "./output/03_st_paths.Rdata")
load(file = "./output/05_stv_calculator.Rdata")
```

## Make Some Sample Travel Data

```{r}
sample_person_1 <- nycts_2018_linked |> 
  filter(person_id == "2201" & svyday_id == 2) |> 
  mutate(
    departure_datetime = case_when(
      row_number() == 1 ~ ymd_hms("2018-03-16  6:45:00"), .default = departure_datetime
    ),
    departure_datetime = case_when(
      row_number() == 2 ~ ymd_hms("2018-03-16  16:30:00"), .default = departure_datetime
    ),
    departure_datetime = case_when(
      row_number() == 3 ~ ymd_hms("2018-03-16  17:15:00"), .default = departure_datetime
    ),
    departure_datetime = case_when(
      row_number() == 4 ~ ymd_hms("2018-03-16  18:25:00"), .default = departure_datetime
    ),
    departure_datetime = case_when(
      row_number() == 5 ~ ymd_hms("2018-03-16  19:55:00"), .default = departure_datetime
    ),
    arrival_datetime = case_when(
      row_number() == 1 ~ ymd_hms("2018-03-16  7:00:00"), .default = arrival_datetime
    ),
    arrival_datetime = case_when(
      row_number() == 2 ~ ymd_hms("2018-03-16  16:45:00"), .default = arrival_datetime
    ),
    arrival_datetime = case_when(
      row_number() == 3 ~ ymd_hms("2018-03-16  17:25:00"), .default = arrival_datetime
    ),
    arrival_datetime = case_when(
      row_number() == 4 ~ ymd_hms("2018-03-16  18:35:00"), .default = arrival_datetime
    ),
    arrival_datetime = case_when(
      row_number() == 5 ~ ymd_hms("2018-03-16  20:00:00"), .default = arrival_datetime
    )
  ) |> 
  mutate(
    person_id = 1, svyday_id = 1,
    trip_id = paste0(person_id, svyday_id, "00", row_number())
  )
```

# Get Episodes

```{r}
sample_episodes <- episode_builder(sample_person_1)
```

don't know what is causing the parse warning, something with lubridate in the case_when?

## Person Weights

add person weights

```{r}
sample_episodes <- sample_episodes |> 
  mutate(per_weight = 1)
```

episode tables

```{r}
activites_short_names <- c(
  "Initial Stay" = "Initial Stay",
  "Transit" = "Transit",
  "Work" = "Work",
  "Walk" = "Walk",
  "Home" = "Home",
  "Restaurant (dine-in or take-out)" = "Restaurant",
  "Friend/Family member’s house" = "Friend/Family",
  "Last Stay" = "Last Stay",
  "K-12 School or daycare" = "Education",
  "Shopping" = "Shopping",
  "Recreation/Entertainment/Exercise (e.g., movies, theater, park)" = "Recreation",
  "Errands or personal business (e.g. bank, post office)" = "Errands",
  "Other" = "Other",
  "Car" = "Car",
  "Work-related place (e.g., meeting, second job, delivery)" = "Work-related",
  "Missing" = "Missing",
  "Religious (e.g., church, synagogue)" = "Religious",
  "Airport" = "Airport",
  "Doctor/Medical Service/Hospital" = "Medical",
  "College/University" = "University",
  "Bicycle" = "Bicycle"
) |> 
  enframe() |> 
  #select(2:1) |> 
  deframe()


trip_table <- sample_episodes |> select(
  person_id,
  #svyday_id,
  trip_id,
  trip_mode,
  departure_time = start_datetime,
  arrival_time = end_datetime,
  origin_type = start_activity_type,
  dest_type = end_activity_type,
  #origin_bg = start_bg,
  #dest_bg = end_bg
) |> 
  drop_na() |> 
  mutate(
    departure_time = hms::as_hms(departure_time),
    arrival_time = hms::as_hms(arrival_time),
    origin_type = activites_short_names[origin_type],
         dest_type = activites_short_names[dest_type])



episode_table <- sample_episodes |> select(
  person_id,
  episode_id,
  activity_class,
  activity_desc,
  #svyday_id,
  #trip_id,
  #trip_mode,
  start_time = start_datetime,
  end_time = end_datetime,
  start_activity_type,
  end_activity_type,
  #start_bg,
  #end_bg
) |> 
  mutate(
    start_time = hms::as_hms(start_time),
    end_time = hms::as_hms(end_time),
    activity_desc = activites_short_names[activity_desc],
    start_activity_type = activites_short_names[start_activity_type],
    end_activity_type = activites_short_names[end_activity_type])



intermediate_table <- episode_table |>
  mutate(
    #person_id = case_when(row_number() %% 2 == 1 ~ NA, .default = person_id), 
    start_time = case_when(row_number() %% 2 == 1 ~ NA_character_, .default = as.character(start_time)), 
    end_time = case_when(row_number() %% 2 == 1 ~ NA_character_, .default = as.character(end_time)),
    #activity_class = case_when(row_number() %% 2 == 1 ~ NA_character_, .default = activity_class),
    activity_desc = case_when(row_number() %% 2 == 1 ~ NA_character_, .default = activity_desc),
    start_activity_type = case_when(row_number() %% 2 == 1 ~ NA_character_, .default = start_activity_type),
    end_activity_type = case_when(row_number() %% 2 == 1 ~ NA_character_, .default = end_activity_type),
    )

trip_table |> tt() |>  style_tt(fontsize = .75) |> theme_tt("spacing")
intermediate_table |> tt() |> format_tt(replace = "-") |> style_tt(fontsize = .75) |> theme_tt("spacing")
episode_table |> tt() |>  style_tt(fontsize = .75) |> theme_tt("spacing")
```

# Make Space-time Paths

get census zone centroid point coordinates

```{r}
census_bg_2010_centroids <- census_bg_2010 |> 
  st_centroid() |> 
  mutate(
    centroid_x = sf::st_coordinates(geometry)[,1],
    centroid_y = sf::st_coordinates(geometry)[,2]
  ) |> 
  select(BlockGroup_GEOID, centroid_x, centroid_y, geometry)
```

build paths

```{r}
sample_st_paths <- st_path_builder(sample_episodes, centroids = census_bg_2010_centroids)
```

```{r}
#st_write(st_paths |> mutate(activity_duration = as.numeric(activity_duration)), "./arcgis_pro/st_paths_test_2201.gpkg")
```

## Filter Paths

```{r}
# get activities
flex_activities <- c(
  "Restaurant (dine-in or take-out)",
  "Recreation/Entertainment/Exercise (e.g., movies, theater, park)",
  "Shopping"
)
```

```{r}
# extract flexible activity episodes
sample_flex_segments <- sample_st_paths |> 
  #filter(person_id == 5101) |> 
  group_by(person_id, timeblock_id) |> 
  #filter(
  #  valid_stay_location != "no" & 
  #    valid_stay_trip_sequence != "no" & 
  #    valid_svyday_sequence != "no") |> 
  mutate(
    # classify valid episode chains
    valid_cones = case_when(
      # is a flexible activity
      activity_desc %in% flex_activities & 
        # bounded by a trip to the activity that occurs immediately prior to the activity
        # non-nyc will be omitted due to missing line segments in workbook 3
        (!is.na(lag(trip_id)) & start_datetime == lag(end_datetime)) &
        # bounded by a trip from the activity that occurs immediately after to the activity
        (!is.na(lead(trip_id)) & end_datetime == lead(start_datetime)) ~ "yes",
      .default = "no"
        )
  ) |> 
  
  # keep only valid cases
  filter(
    valid_cones == "yes" |
      # previous episode
      (lag(valid_cones) == "yes") |
      
      # next episode
      (lead(valid_cones) == "yes")
    ) |> 
  group_by(person_id) |> 
  
  # create an id for the flexible sequences
  mutate(flexseq_id = cumsum(start_datetime != lag(end_datetime, default = first(end_datetime))) + 1) |> 
  
  # get geometry as a column from lines
  mutate(
    start_x =  sf::st_coordinates(lwgeom::st_startpoint(geometry))[,1],
    start_y = sf::st_coordinates(lwgeom::st_startpoint(geometry))[,2],
    end_x = sf::st_coordinates(lwgeom::st_endpoint(geometry))[,1],
    end_y = sf::st_coordinates(lwgeom::st_endpoint(geometry))[,2]) |> 
  st_drop_geometry() |>
  
  # summarize down to one row for each flex segment
  group_by(person_id, timeblock_id, flexseq_id) |>
  summarize(
    num_episodes = n(),
    start_trip_mode = first(trip_mode),
    end_trip_mode = last(trip_mode),
    start_datetime = first(start_datetime),
    end_datetime = last(end_datetime),
    time_window = sum(activity_duration),
    per_weight = first(per_weight),
    start_x = first(start_x),
    start_y = first(start_y),
    end_x = last(end_x),
    end_y = last(end_y)
  ) |> 
  ungroup() |> 
  tidyr::unite(col = "id", c(person_id, timeblock_id, flexseq_id))
```

add another flex segment for comparison

```{r}
sample_flex_segments <- sample_flex_segments |> 
  mutate(count = 2) |> 
  tidyr::uncount(count, .id = "count_id") |> 
  mutate(
    start_x = case_when(count_id == 2 ~ 591000, .default = start_x),
    end_x = case_when(count_id == 2 ~ 591000, .default = end_x),
    start_y = case_when(count_id == 2 ~ 4513500, .default = start_y),
    end_y = case_when(count_id == 2 ~ 4513500, .default = end_y),
    start_datetime = case_when(count_id == 2 ~ start_datetime + minutes(120), .default = start_datetime),
    end_datetime = case_when(count_id == 2 ~ start_datetime + minutes(100), .default = end_datetime),
    id = case_when(count_id == 2 ~ "2_1_1", .default = id),
    time_window = difftime(end_datetime, start_datetime, units = "mins")
  )
```

# Routing

```{r}
r5_path_forward <- dir_create("./r5_graph_forward")
r5_graph_forward <- r5r::build_network(data_path = r5_path_forward)

r5_path_reverse <- fs::dir_create("./r5_graph_reverse")
r5_graph_reverse <- r5r::build_network(data_path = r5_path_reverse)
```

## Prepare 10m study area grid

```{r}
cell_size = 10

study_area_bbox <- boundaries_boroughs |> 
  filter(NAMELSAD == "New York County") |> 
  st_buffer(1000) |> 
  st_bbox()

study_area_boundaries <- boundaries_boroughs |> 
  st_crop(study_area_bbox) |> 
  st_collection_extract("POLYGON")

study_area_raster <- study_area_boundaries |>
  select(geometry) |> 
  st_rasterize(
    template = st_as_stars(
      study_area_bbox,
      dx = cell_size,
      dy = cell_size,
      pretty = TRUE,
      values = NA_real_))

study_area_points <- study_area_raster |> 
  st_xy2sfc(as_points = TRUE) |> 
  st_as_sf() |> 
  select(-ID) |> 
  mutate(id = as.character(row_number()))

study_area_polygons <- study_area_raster |> 
  st_xy2sfc(as_points = FALSE) |> 
  st_as_sf() |> 
  select(-ID) |> 
  st_join(study_area_points)
```

## Get Opportunities in Intervening Grid Cells

```{r}
overture_flexthemes <- c(
  "active_life",
  "arts_and_entertainment",
  "attractions_and_activities",
  "eat_and_drink",
  "retail"
)

# count opportunities by cell
nyc_places_flex <- nyc_places |> 
  rename(poi_id = id) |> 
  filter(theme_primary %in% overture_flexthemes) |> 
  st_transform(crs = 26918) |> 
  st_join(study_area_polygons) |>
  drop_na(id) |> 
  st_drop_geometry() |> 
  group_by(id) |> 
  summarize(opp_weight = n())

# join back to study area points
study_area_points <- study_area_points |> 
  left_join(nyc_places_flex, by = "id") |> 
  mutate(opp_weight = replace_na(opp_weight, 0))

opportunities_raster <- study_area_points |> 
  select(-id) |> 
  filter(opp_weight > 0) |> 
  st_as_stars() |> 
  st_sfc2xy()
```

## Define Time Reversal Function

```{r}
#| label: define reverse time and datetime functions
# define reverse time function
reverse_time <- function(forward_time){
  hh_rev = str_pad(abs(hour(forward_time) - 23), width = 2, side = "left", pad = "0")
  mm_rev = str_pad(abs(minute(forward_time) - 59), width = 2, side = "left", pad = "0")
  ss_rev = str_pad(abs(second(forward_time) - 59), width = 2, side = "left", pad = "0")
  time_rev = paste(hh_rev, mm_rev, ss_rev, sep = ":")
  return(time_rev)
}

# define reverse datetime function
reverse_datetime <- function(forward_datetime) {
  forward_date = as_date(forward_datetime)
  time_rev = reverse_time(forward_datetime)
  datetime_rev = ymd_hms(paste(forward_date, time_rev, sep = " "))
  return(datetime_rev)
}
```

## Calculate Space-time Voxels

```{r}
stv_sample_path <- fs::dir_create("./stvs_sample_10m")
time_unit = 5

tictoc::tic()
sample_flex_segments |> 
  drop_na() |> 
  filter(start_trip_mode != "OTHER" & end_trip_mode != "OTHER") |> 
  group_by(id) |> 
  group_walk(~ stv_calculator(
    data = .x, 
    group_keys = .y, 
    intermediate_points = study_area_points,
    time_unit = time_unit, # minutes
    output_path = stv_sample_path))
  
tictoc::toc()
```

# Get Stars Objects
## STVs

```{r}
stvs_sample <- arrow::open_dataset("./stvs_sample_10m") |> 
  collect() |> 
  mutate(
    z = as.numeric(difftime(
      t, 
      lubridate::floor_date(t, unit = "day"),
      units = "mins")
    )) |> 
  select(x, y, level = z, person_id, dwell_time) |> 
  arrange(x, y, level)

stvs_sample_person_1_stars <- stvs_sample |> 
  filter(person_id == 1) |> 
  st_as_stars(dims = c("x", "y", "level")) |> 
  st_set_crs(26918)

stvs_sample_person_2_stars <- stvs_sample |> 
  filter(person_id == 2) |> 
  st_as_stars(dims = c("x", "y", "level")) |> 
  st_set_crs(26918)
```

```{r}
#| eval: false
write_mdim(stvs_sample_person_1_stars, "./arcgis_pro/stvs_sample_person_1_stars.nc")
write_mdim(stvs_sample_person_2_stars, "./arcgis_pro/stvs_sample_person_2_stars.nc")
```

## PPA Rasters

```{r}
pasta_sample_person_1 <- arrow::open_dataset("./stvs_sample_10m") |> 
  filter(person_id == 1) |> 
  collect() |> 
  group_by(x, y) |> 
  summarize(dwell_time = sum(dwell_time)) |> 
  select(x, y, dwell_time) |> 
  arrange(x, y) |> 
  st_as_stars(dims = c("x", "y")) |> 
  st_set_crs(26918)

pasta_sample_person_2 <- arrow::open_dataset("./stvs_sample_10m") |> 
  filter(person_id == 2) |> 
  collect() |> 
  group_by(x, y) |> 
  summarize(dwell_time = sum(dwell_time)) |> 
  select(x, y, dwell_time) |> 
  arrange(x, y) |> 
  st_as_stars(dims = c("x", "y")) |> 
  st_set_crs(26918)

pasta_sample <- arrow::open_dataset("./stvs_sample_10m") |> 
  collect() |> 
  group_by(x, y) |> 
  summarize(dwell_time = sum(dwell_time)) |> 
  select(x, y, dwell_time) |> 
  arrange(x, y) |> 
  st_as_stars(dims = c("x", "y")) |> 
  st_set_crs(26918)
```

```{r}
#| eval: false
stars::write_stars(pasta_sample_person_1, "./arcgis_pro/pasta_sample_person_1.tif")
stars::write_stars(pasta_sample_person_2, "./arcgis_pro/pasta_sample_person_2.tif")
stars::write_stars(pasta_sample, "./arcgis_pro/pasta_sample.tif")
```

# Save Output

```{r}
save(trip_table, intermediate_table, episode_table, file = "./output/09_samples.Rdata")
```

